---
output: md_document
---
## Analysis of economic and public health consequences of severe weather events in US

Final project for Reproducible Research course on Coursera (5th course of the data analysis specialization).

### Synopsis

This project analyses the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

According to the findings, Tornado, Wind and High Water are responsible for the biggest economic losses. In absolute dollars, the property damage is bigger than the crops damage across all events except Heat and Tempreture events. Tornados also have the most dramatic negative impact on public health: they are responsible for twice as many injuries as all other weather conditions combined.

```{r, echo = FALSE}
setwd("C:/Users/a/Desktop/DS/5 Reproducible Research/assignment2")
```
### Data Processing
Download storm.RData file with variable *storm* if we have it in the working directory. If not - loading and reading the data in *storm* variable. Notice that read.csv function allows us to read a compressed file.
```{r, cache = TRUE}
storm <- 0
# check if storm variable already exists
if (file.exists("storm.RData")){
      load("storm.RData")
} else {
      if (!file.exists("storm.bz2")) {
            url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
            download.file(url, destfile= "storm.bz2")
      }
      storm <- read.csv("storm.bz2")
      save(storm, file="storm.RData")
}
```

Remove all unwanted columns so that we have only variables related to public health, economic losses and names of the events.
```{r, cache = TRUE}
subData <- storm[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "CROPDMG")]
```

Also delete rows with "?" and summaries in the event name column "EVTYPE".
```{r, cache = TRUE}
# Delete the row with "?" in the EVTYPE column
subData <- subData[!(subData$EVTYPE == "?"),]
# Remove summaries
subData <- subData[!(grepl('Summary', subData$EVTYPE, ignore.case = TRUE)),]
```

Now let's unite 985 unique values of events into meaningfull groups.
```{r, cache = TRUE}
subData$EVTYPE <- as.character(subData$EVTYPE)
subData[grepl("cold|cool|ice|icy|frost|freeze|snow|winter|wintry|wintery|blizzard|
              chill|freezing|glaze|sleet", 
              subData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "Cold Weather"

subData[grepl("warmth|DRIEST MONTH|warm|heat|dry|hot|drought|thermia|temperature record|
            record temperature|record high|temperature", 
            subData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "Heat and Temperature"

subData[grepl("wind|storm|wnd|waterspouts|waterspouts/", 
              subData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "Wind"

subData[grepl("precipitation|MIXED PRECIP|rain|hail|drizzle|LIGNTNING|HEAVY SHOWER |
            LIGHTNING|LIGHTING|TSTM|HEAVY SHOWER[S]|HEAVY PRECIPATATION|
            wet|percip|burst|[fv]og|wall cloud", 
            subData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "Precipitation and Thunderstorm"

subData[grepl("hurricane|typhoon|tornado|TORNDAO|funnel|whirlwind", 
              subData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "Tornado"

subData[grepl("dust|GUSTNADO|saharan|landspout", subData$EVTYPE,
              ignore.case = TRUE), "EVTYPE"] <- "Dust"

subData[grepl("seas|high water|fld|fldg|dam|water|SURGE|SEICHE|SWELLS|tide|tsunami|wave|
      current|surf|marine|drowning|flood|DAM FAILURE|WATERSPOUT|WATER SPOUT|WAYTERSPOUT", 
              subData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "High Water"

subData[grepl("ash|smoke|volcanic|fire", 
              subData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "Fire and Volcanic Activity"

subData[grepl("slide|erosion|EROSIN|landslump", 
              subData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "Ground Movements"

subData[!grepl(paste("Heat and Temperature|Wind|Tornado|Dust|Precipitation and Thunderstorm|Cold .Weather",
                     "|High Water|Ground Movements|Fire and Volcanic Activity", sep = ""), 
            subData$EVTYPE), "EVTYPE"] <- "Other"
```

### Data Analysis
First, let's explore which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health. In order to do that, we aggregate and sort data about injuries and fatalities by the new categories. The end result is the variable "HHarmSubData" which stands for human harm sub data.

```{r, message = F}
stopifnot(require("dplyr"))
ArrangedSubData <- aggregate(cbind(INJURIES, FATALITIES) ~ EVTYPE, data = subData, FUN = sum)
HHarmSubData <- ArrangedSubData[order(ArrangedSubData$INJURIES, decreasing = T),]
HHarmSubData$EVTYPE <- factor(HHarmSubData$EVTYPE, levels=(HHarmSubData$EVTYPE))
```

Now the second question. Across the United States, which types of events have the greatest economic consequences? We go through the same set of commands to create aggregated, sorted economic harm data EHarmSubData.
```{r}
EHarmSubData <- aggregate(cbind(PROPDMG, CROPDMG) ~ EVTYPE, data = subData, FUN = sum)
EHarmSubData <- EHarmSubData[order(EHarmSubData$PROPDMG, decreasing = T),]
EHarmSubData$EVTYPE <- factor(EHarmSubData$EVTYPE, levels=(EHarmSubData$EVTYPE))
```

### Results
Lets look at at the resulting data.
```{r}
HHarmSubData
EHarmSubData
```

Let's look at the graphical representation of the results.
```{r}
# Load required packages
stopifnot(require("ggplot2")) # package for the graphs
library(reshape2)  # package for melt function

HHarmSubData %>% melt(id.vars = "EVTYPE", variable.name = "Event", value.name = "Number_of_People") %>%
      ggplot(aes(x = Event, y = Number_of_People, col = Event)) +
      geom_bar(stat = "identity", fill = "white") + 
      facet_wrap( ~ EVTYPE, nrow = 1) + theme(axis.text.x = element_blank(), 
      strip.text.x = element_text(angle = 90, size=8, hjust = .5, vjust = .5)) +
      ggtitle("Number of people injured and killed by weather events")
```

Injuries from tornados are overwhelmingly high. Let's calculate the ratio of number of people injured during tornados to the number of people injured during all other weather events.
```{r}
HHarmSubData[1,2]/sum(HHarmSubData[2:7,2])
```
As we can see, the number of people injured during tornados is twice as high as the number of people injured during all other weather conditions combined. Number of fatalities is also the biggest between all other groups of events.

Now to the economic losses.
```{r}
EHarmSubData %>% melt(id.vars = "EVTYPE", variable.name = "Event", value.name = "Economic_Loss") %>%
      ggplot(aes(x = Event, y = Economic_Loss/1e+6, col = Event)) +
      geom_bar(stat = "identity", fill = "white") + 
      facet_wrap( ~ EVTYPE, nrow = 1) + theme(axis.text.x = element_blank(), 
      strip.text.x = element_text(angle = 90, size=8, hjust = .5, vjust = .5)) +
      ggtitle("Economic losses by weather events") + ylab("Economic Losses in millions")

EHarmSubData[,"PROPDMG"] > EHarmSubData[,"CROPDMG"]

```


Three most damaging groups of events in terms of economic losses are Tornado, Wind and High Water respectively. The most damaging events for the crops are precipitation and thunderstorm And in terms of costs, property damage is much more expensive than the damage to the crop in all cases except Heat and Temperature events.